plyranges: A grammar of genomic data transformation

### 2018-09-20 paper.Rmd
use latex instead of rmd for figure, output data and code.

@@

```{r olaps-fig, echo = FALSE, out.width="\\textwidth", fig.cap="Illustration of the three overlap join operators. Each join takes two GRanges objects, \\textit{x} and \\textit{y} as input. A 'Hits' object for the join is computed which consists of two components. The first component contains the indices of the ranges in \\textit{x} that have been overlapped (the rectangles of \\textit{x} that cross the orange lines). The second component consists of the indices of the ranges in \\textit{y} that overlap the ranges in \\textit{x}. In this case a range in \\textit{y} overlaps the ranges in \\textit{x} three times, so the index is repeated three times. Te resulting 'Hits' object is used to modify \\textit{x} by where it was 'hit' by \\textit{y}. This principle of is applied generally in the \\texttt{plyranges} DSL for both overlap and nearest neighbour operations. The join semantics alter what is returned: \\textbf{A}: for an \\textbf{inner} join the \textit{x} ranges that are overlapped by \\textit{y} are returned. \\textbf{B}  For an \\textbf{intersect} join the intersection is taken between the overlapped \\textit{x} ranges and the \\textit{y} ranges. \\textbf{C} For the left join \\textbf{left} join all \\textit{x} ranges are returned regardless of whether they are overlapped by \\textit{y}."}
knitr::include_graphics("diagrams/olap-fig.png")
```

@@

Since the GRanges object is a tabular data structure, our grammar includes
operators to filter, sort and aggregate by columns in a GRanges. These 
operations can be performed over partitions formed
using the *group_by* modifier. Together with our algebra for arithmetic and
merging, these operations conform to the semantics and syntax of the 
`dplyr` grammar. Consequently, this results

@@

Here we provide illustrative examples of using the `plyranges` DSL to show
how our grammar could be integrated into genomic data workflows. We also
highlight how interoperability with existing Bioconductor infrastructure, 
enables easy access to public datasets and methods for analysis and 
visualization.

@@

```{r}
chr10_scores <- bw_file %>%
  read_bigwig(overlap_ranges = chr10_ranges) %>%
  set_genome_info(genome = "hg19")
```

@@

```{r}
all_peaks <- chr10_scores %>% 
  filter(score > 8) %>% 
  reduce_ranges(score = max(score))
```

@@

```{r peak_region}
peak_region <- chr10_scores %>%
  join_overlap_inner(chr10_max_score_region)

```

@@

```{r, cache = TRUE}
alignments <- h1_bam_sorted %>% 
  read_bam() %>%
  filter_by_overlaps(locations) %>%
  select(seq) %>%
  mutate(score = as.numeric(letterFrequency(seq, "GC", as.prob = TRUE)))
```

>>>>

\begin{figure}
\includegraphics[width=\textwidth]{diagrams/olap-fig}
\caption{Illustration of the three overlap join operators. Each join takes two 
GRanges objects, \\textit{x} and \\textit{y} as input. A 'Hits' object for the 
join is computed which consists of two components. The first component contains 
the indices of the ranges in \\textit{x} that have been overlapped 
(the rectangles of \\textit{x} that cross the orange lines). The second 
component consists of the indices of the ranges in \\textit{y} that overlap 
the ranges in \\textit{x}. In this case a range in \\textit{y} overlaps the 
ranges in \\textit{x} three times, so the index is repeated three times. The 
resulting 'Hits' object is used to modify \\textit{x} by where it was 'hit' by
\\textit{y} and merge all metadata columns from \\textit{x} and \\textit{y} 
based on the indices contained in the 'Hits' object. This procedure is applied 
generally in the \\texttt{plyranges} DSL for both overlap and nearest neighbour
operations. The join semantics alter what is returned: \\textbf{A}: for an 
\\textbf{inner} join the \\textit{x} ranges that are overlapped by \\textit{y} 
are returned. The returned ranges also include the metadata from the \\textit{y}
range that overlapped the three \\textit{x} ranges. \\textbf{B} An 
\\textbf{intersect} join is identical to an inner join except that the 
intersection is taken between the overlapped \\textit{x} ranges and the 
\\textit{y} ranges. \\textbf{C} For the left join \\textbf{left} join all 
\\textit{x} ranges are returned regardless of whether they are overlapped by 
\\textit{y}. In this case the resulting third range  of the join would have 
missing values on metadata columns that came from \\textit{y}.}
\label{fig:olaps-fig}
\end{figure}

@@

@@

Since the GRanges object is a tabular data structure, our grammar includes
operators to filter, sort and aggregate by columns in a GRanges. These 
operations can be performed over partitions formed
using the *group_by* modifier. Together with our algebra for arithmetic and
merging, these operations conform to the semantics and syntax of the 
`dplyr` grammar. Consequently, `plyranges` code is generally more compact
than the equivalent `GenomicRanges` code (figure \ref{code-fig}).

@@

Here we provide illustrative examples of using the `plyranges` DSL to show
how our grammar could be integrated into genomic data workflows. As we
construct the workflows we show the data output intermittantly to assist
the reader in understanding the pipeline steps. The workflows
highlight how interoperability with existing Bioconductor infrastructure, 
enables easy access to public datasets and methods for analysis and 
visualization.

@@

```{r}
chr10_scores <- bw_file %>%
  read_bigwig(overlap_ranges = chr10_ranges) %>%
  set_genome_info(genome = "hg19")
chr10_scores
```

@@

```{r}
all_peaks <- chr10_scores %>% 
  filter(score > 8) %>% 
  reduce_ranges(score = max(score))
all_peaks
```

@@

```{r peak_region}
peak_region <- chr10_scores %>%
  join_overlap_inner(chr10_max_score_region)
peak_region
```

@@

```{r, cache = TRUE}
alignments <- h1_bam_sorted %>% 
  read_bam() %>%
  filter_by_overlaps(locations) %>%
  select(seq) %>%
  mutate(score = as.numeric(letterFrequency(seq, "GC", as.prob = TRUE)))
alignments
```
